---
title: "Cluster Annotation with SingleR"
author: "Eileen Owens"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      cache = TRUE,
                      warning = FALSE,
                      keep_md = TRUE)
```

# Introduction
The purpose of this script is to compare single-cell expression profiles with previously annotated reference datasets using SingleR. This script has been adapted from the Seurat (https://satijalab.org/seurat/) and SingleR (https://bioconductor.org/packages/devel/bioc/vignettes/SingleR/inst/doc/SingleR.html) documentation.

# Software packages
```{r software-load, results="hide", message=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(SingleR)
library(celldex) # To install: BiocManager::install("celldex")
library(data.table)
```

# Data
## Working directory
```{r setwd}
setwd("C:/Users/edlarsen/Documents/240828_scRNAseq/Cluster_Annotation")
```

## Filtered, normalized, and clustered Seurat object(s)
```{r data-import}
ln_seurat <- readRDS(file = "../Normalization_and_Clustering/LN_NormalizedAndClustered.RData")
thym_seurat <- readRDS(file = "../Normalization_and_Clustering/THYM_NormalizedAndClustered.RData")
```

# Differentially Expressed Features
By default, identifies all positive and negative markers of a single cluster compared to all other cells.
```{r FindAllMarkers, message=FALSE}
## Lymph node
ln_seurat <- FindClusters(ln_seurat, resolution = 0.1)
ln_seurat <- RunUMAP(ln_seurat, dims = 1:10)
ln.markers <- FindAllMarkers(ln_seurat, only.pos=TRUE)
DimPlot(ln_seurat,
        reduction = "umap",
                   label = TRUE,
                   label.size = 6) + 
  plot_annotation(title = "Canine Lymph Node, Resolution: 0.1", theme = theme(plot.title = element_text(hjust = 0.5, size = 20)))

## Thymus
thym_seurat <- FindClusters(thym_seurat, resolution = 0.2)
thym_seurat <- RunUMAP(thym_seurat, dims = 1:10)
thym.markers <- FindAllMarkers(thym_seurat, only.pos=TRUE)
thym_seurat <- RunUMAP(thym_seurat, 
                           dims = 1:10,
                           n.neighbors = 50, # default is 30
                           min.dist = 0.5) # default is 0.3
DimPlot(thym_seurat,
        reduction = "umap",
                   label = TRUE,
                   label.size = 6) + 
  plot_annotation(title = "Canine Thymus, Resolution: 0.2, \nn.neighbors = 50, min.dist = 0.5", theme = theme(plot.title = element_text(hjust = 0.5, size = 20)))
```

# SingleR
SingleR is a computational method using reference transcriptomic datasets of pure cell types for unbiased cell type recognition in single-cell RNA-seq data. The SingleR method assigns labels to cells based on the reference sample with the highest Spearman rank correlations, using only the marker genes between pairs of labels to focus on the relevant differences between cell types. It also performs a fine-tuning step for each cell, where the correlations are recomputed with just the marker genes for the top-scoring labels. This aims to resolve any ambiguity between those labels by removing noise from irrelevant markers for other labels.

*celldex* contains a number of curated reference datasets, mostly assembled from bulk RNA-seq or microarray data of sorted cell types. Here, a  will be used.

## Blueprint and ENCODE Data
Using a reference constructed from Blueprint and ENCODE data (Martens and Stunnenberg 2013 and ENCODE Project Consortium 2012).
```{r BlueprintEncode}
# pull Blueprint and ENCODE reference data from celldex
ref_bpe <- celldex::BlueprintEncodeData()

# See what cell types are included in the reference and subset to include only those relevant to the single-cell data
unique(ref_bpe$label.main)
ref_bpe <- ref_bpe[,grepl("Neutrophils|Macrophages|Monocytes|Erythrocytes|HSC|Endothelial cells|CD4+ T-cells|DC|CD8+ T-cells|Eosinophils|NK cells|B-cells|Fibroblasts", ref_bpe$label.main)]

# extract normalized counts from Seurat object
norm_counts_ln <- LayerData(ln_seurat, assay = "RNA", layer = "data")
norm_counts_thym <- LayerData(thym_seurat, assay = "RNA", layer = "data")

# use singleR function to annotate cells with labels from reference
bpe_ln <- SingleR(test=norm_counts_ln, ref=ref_bpe, labels=ref_bpe$label.main)
bpe_thym <- SingleR(test=norm_counts_thym, ref=ref_bpe, labels=ref_bpe$label.main)
```

### Inspect quality of predictions
#### Heatmap
Each column in the heatmap represents a cell. Reference labels on y-axis. We expect cells with a high score for T cells and low for the other clusters to have the final label of T cells.
If prediction for a certain cell is ambiguous, it will have high scores (yellow) for many cell identities.
```{r singleR-bpe-heatmap-ln}
plotScoreHeatmap(bpe_ln)
```

```{r singleR-bpe-heatmap-thym}
plotScoreHeatmap(bpe_thym)
```

#### Delta distribution
Shows the distribution of the gap between the score for the assigned label and the score of the remaining label(s) across cells assigned to each reference label. We want high deltas, as they give us more confidence in the assigned label.

By default, SingleR will always assign a label to every cell, resulting in incorrect label assignment if the cell's true label isn't in the reference data set. To mitigate this, a "delta" value is calculated for each cell, representing the difference between the score for the assigned label and the median score across all labels. If the delta is small, this indicates that the cell matches all labels with the same confidence, so the assigned label is not very meaningful. SingleR can then discard cells with low delta values caused by ambiguous assignments with closely related reference labels and incorrect assignments that match poorly to all reference labels. 

The 'cleaner' results are found in the pruned_labels column, cells that have deltas that were too small are assigned an NA value:
```{r pruned-cells-bpe}
# The number of cells assigned NA values (deltas too small) will be tallied in the TRUE column
paste("Pruned cells - Lymph node:", summary(is.na(bpe_ln$pruned.labels)))
paste("Pruned cells - Thymus:", summary(is.na(bpe_thym$pruned.labels)))
```

```{r bpe-ln-delta-plot, fig.height = 10, fig.width = 8}
plotDeltaDistribution(bpe_ln, ncol = 4, dots.on.top = FALSE) + ggtitle("Canine Lymph Node Delta Values for Blueprint and ENCODE Annotations")
```

```{r bpe-thym-delta-plot, fig.height = 10, fig.width = 8}
plotDeltaDistribution(bpe_thym, ncol = 4, dots.on.top = FALSE) + ggtitle("Canine Thymus Delta Values for Blueprint and ENCODE Annotations")
```

## Human Primary Cell Atlas Data
```{r HPCAtlas}
# pull reference data from celldex
ref_hpca <- celldex::HumanPrimaryCellAtlasData()

# See what cell types are included in the reference and subset to include only those relevant to the single-cell data
unique(ref_hpca$label.main)
ref_hpca <- ref_hpca[,grepl("DC|BM & Prog.|Embryonic_stem_cells|Endothelial_cells|Smooth_muscle_cells|MSC|Pro-Myelocyte|BM|HSC_CD34+|HSC_-G-CSF|GMP|CMP|MEP|Pro-B_cell_CD34-|Pro-B_cell_CD34+|B_cell|Neutrophils|T_cells|Monocyte|Erythroblast|Macrophage|NK_cell|Platelets|Myelocyte", ref_hpca$label.main)]

# use singleR function to annotate cells with labels from reference
hpca_ln <- SingleR(test=norm_counts_ln, ref=ref_hpca, labels=ref_hpca$label.main)
hpca_thym <- SingleR(test=norm_counts_thym, ref=ref_hpca, labels=ref_hpca$label.main)
```

### Inspect quality of predictions
#### Heatmap
```{r singleR-hpca-heatmap-ln}
plotScoreHeatmap(hpca_ln)
```

```{r singleR-hpca-heatmap-thym}
plotScoreHeatmap(hpca_thym)
```

#### Delta distribution
```{r pruned-cells-hpca}
# The number of cells assigned NA values (deltas too small) will be tallied in the TRUE column
paste("Pruned cells - Lymph node:", summary(is.na(hpca_ln$pruned.labels)))
paste("Pruned cells - Thymus:", summary(is.na(hpca_thym$pruned.labels)))
```

```{r hpca-ln-delta-plot, fig.height = 10, fig.width = 8}
plotDeltaDistribution(hpca_ln, ncol = 4, dots.on.top = FALSE) + ggtitle("Canine Lymph Node Delta Values for Human Primary Cell Atlas Annotations")
```

```{r hpca-thym-delta-plot, fig.height = 10, fig.width = 8, warning = FALSE}
plotDeltaDistribution(hpca_thym, ncol = 5, dots.on.top = FALSE) + ggtitle("Canine Thymus Delta Values for Human Primary Cell Atlas Annotations")
```

### Using fine labels
```{r HPCAtlas-FineLabels}
# pull reference data from celldex
ref_hpca_fine <- celldex::HumanPrimaryCellAtlasData()

# extract normalized counts from Seurat object
norm_counts_ln <- LayerData(ln_seurat, assay = "RNA", layer = "data")
norm_counts_thym <- LayerData(thym_seurat, assay = "RNA", layer = "data")

# use singleR function to annotate cells with labels from reference
hpca_ln_fine <- SingleR(test=norm_counts_ln, ref=ref_hpca_fine, labels=ref_hpca_fine$label.fine)
hpca_thym_fine <- SingleR(test=norm_counts_thym, ref=ref_hpca_fine, labels=ref_hpca_fine$label.fine)
```

### Inspect quality of predictions
#### Heatmap
```{r singleR-hpcafine-heatmap-ln, fig.width = 16}
plotScoreHeatmap(hpca_ln_fine)
```

```{r singleR-hpcafine-heatmap-thym, fig.width = 16}
plotScoreHeatmap(hpca_thym_fine)
```

#### Delta distribution
```{r pruned-cells-hpcafine}
# The number of cells assigned NA values (deltas too small) will be tallied in the TRUE column
paste("Pruned cells - Lymph node:", summary(is.na(hpca_ln_fine$pruned.labels)))
paste("Pruned cells - Thymus:", summary(is.na(hpca_thym_fine$pruned.labels)))
```

```{r hpcafine-ln-delta-plot, fig.height = 20, fig.width = 14}
plotDeltaDistribution(hpca_ln_fine, ncol = 6, dots.on.top = FALSE) + ggtitle("Canine Lymph Node Delta Values for Human Primary Cell Atlas Annotations")
```

```{r hpcafine-thym-delta-plot, fig.height = 20, fig.width = 16, warning = FALSE}
plotDeltaDistribution(hpca_thym_fine, ncol = 5, dots.on.top = FALSE) + ggtitle("Canine Thymus Delta Values for Human Primary Cell Atlas Annotations")
```


# Add SingleR predictions to Seurat object
## Blueprint and ENCODE Data
```{r umap-bpe-ln, fig.height=6}
ln_seurat <- AddMetaData(ln_seurat, bpe_ln$pruned.labels, col.name = "SingleR_BPE")

ln_seurat <- SetIdent(ln_seurat, value = "SingleR_BPE")
DimPlot(ln_seurat, label = T, repel = T, label.size = 3) + NoLegend() + ggtitle("Blueprint and ENCODE Annotations: Canine Lymph Node")
```

```{r umap-bpe-thym, fig.height=6}
thym_seurat <- AddMetaData(thym_seurat, bpe_thym$pruned.labels, col.name = "SingleR_BPE")

thym_seurat <- SetIdent(thym_seurat, value = "SingleR_BPE")
DimPlot(thym_seurat, label = T, repel = T, label.size = 3) + NoLegend() + ggtitle("Blueprint and ENCODE Annotations: Canine Thymus")
```

## Human Primary Cell Atlas Data
```{r umap-hpca-ln, fig.height=6}
ln_seurat <- AddMetaData(ln_seurat, hpca_ln$pruned.labels, col.name = "SingleR_HPCA")

ln_seurat <- SetIdent(ln_seurat, value = "SingleR_HPCA")
DimPlot(ln_seurat, label = T, repel = T, label.size = 3) + NoLegend() + ggtitle("Human Primary Cell Atlas Annotations - Main Labels: \nCanine Lymph Node")
```

```{r umap-hpca-ln-fine, fig.height=20, fig.width = 20}
ln_seurat <- AddMetaData(ln_seurat, hpca_ln_fine$pruned.labels, col.name = "SingleR_HPCA")

ln_seurat <- SetIdent(ln_seurat, value = "SingleR_HPCA")
DimPlot(ln_seurat, label = T, repel = T, label.size = 5) + NoLegend() + ggtitle("Human Primary Cell Atlas Annotations - Fine Labels: Canine Lymph Node")
```

```{r umap-hpca-thym, fig.height=6}
thym_seurat <- AddMetaData(thym_seurat, hpca_thym$pruned.labels, col.name = "SingleR_HPCA")

thym_seurat <- SetIdent(thym_seurat, value = "SingleR_HPCA")
DimPlot(thym_seurat, label = T, repel = T, label.size = 3) + NoLegend() + ggtitle("Human Primary Cell Atlas Annotations - Main Labels: Canine Thymus")
```

```{r umap-hpca-thym-fine, fig.height=10}
thym_seurat <- AddMetaData(thym_seurat, hpca_thym_fine$pruned.labels, col.name = "SingleR_HPCA")

thym_seurat <- SetIdent(thym_seurat, value = "SingleR_HPCA")
DimPlot(thym_seurat, label = T, repel = T, label.size = 4) + NoLegend() + ggtitle("Human Primary Cell Atlas Annotations - Fine Labels: Canine Thymus")
```

# Citations
```{r citations}
sessionInfo()
citation()
```
